---
title: "Getting Started with DICi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DICi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

```{r, eval = FALSE}
remotes::install_github("DoriaXiao/DIC_i")
```

## What is DIC_i?

DIC_i is a parameterization-invariant, plug-in-free version of the Deviance
Information Criterion for Bayesian model comparison. It adds the
variance-based penalty of Gelman et al. (2014b) to the posterior **mean**
deviance, rather than to the unstable plug-in deviance used by the classical
DIC:

$$\mathrm{DIC}_i = E[D(\theta)] + \frac{1}{2} \mathrm{Var}[D(\theta)]$$

where $D(\theta) = -2 \log f_m(y \mid \theta)$ is the marginal deviance.
The penalty $p_V = \frac{1}{2}\mathrm{Var}(D)$ is always non-negative, and
because DIC_i never evaluates the deviance at a point estimate, the whole
criterion is invariant to reparameterization.

This is what distinguishes DIC_i from the related variance-based DIC of
Gelman et al. (2014b), $\mathrm{DIC}_p = D(\bar\theta) + 2 p_V$, which keeps
the plug-in deviance $D(\bar\theta)$ and so remains unstable under
multimodality.

## Quick start

The core function takes an S x N matrix of pointwise marginal
log-likelihoods (rows = posterior draws, columns = observations):

```{r basic}
library(DICi)

# Simulate a log-likelihood matrix (S=1000 draws, N=50 observations)
set.seed(42)
log_lik <- matrix(rnorm(1000 * 50, mean = -2, sd = 0.5), nrow = 1000, ncol = 50)

result <- compute_dic_i(log_lik)
print(result)
```

## Using deviance draws directly

If you only have the joint marginal deviance (e.g., from Mplus output),
pass it directly:

```{r deviance}
deviance_draws <- -2 * rowSums(log_lik)
result2 <- compute_dic_i(deviance_draws = deviance_draws)
print(result2)
```

The results are identical because DIC_i only needs $E[D]$ and $\mathrm{Var}(D)$.

## Comparing models

Fit multiple candidate models, compute DIC_i for each, and compare:

```{r compare}
set.seed(123)
ll_simple  <- matrix(rnorm(1000 * 50, mean = -2.0, sd = 0.4), 1000, 50)
ll_complex <- matrix(rnorm(1000 * 50, mean = -1.9, sd = 0.6), 1000, 50)

fit_simple  <- compute_dic_i(ll_simple)
fit_complex <- compute_dic_i(ll_complex)

compare_dic_i(simple = fit_simple, complex = fit_complex)
```

The model with the lowest DIC_i is preferred. The `delta_dic_i` column shows
the difference from the best model.

## Important: Marginal log-likelihoods required

For latent variable models (factor analysis, mixture models, multilevel
models), the input must be **marginal** log-likelihoods, integrated over the
latent variables. Conditional log-likelihoods (evaluated at specific latent
variable draws) will produce misleading results.

See `vignette("stan-workflow")` for a complete example of fitting a factor
analysis model in Stan and extracting marginal log-likelihoods.

## Next steps

- `vignette("stan-workflow")` --- Full Stan example with sign switching
  demonstration
- `?compute_dic_i` --- Function reference with details on marginal vs.
  conditional likelihoods
- `dici_example()` --- Copy example Stan model and R script to your
  working directory
