---
title: "Stan Workflow: Factor Analysis with Sign Switching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stan Workflow: Factor Analysis with Sign Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette walks through a complete example: simulating data from a
one-factor model, fitting it in Stan with symmetric priors (which causes
sign switching), and showing that DIC_i remains stable while the classical
DIC fails.

To run this example yourself, copy the Stan model and R script to your
working directory:

```{r copy-example}
library(DICi)

# Copy example files to your working directory
dici_example("fa_marginal.stan")
dici_example("example_fa.R")

# Then run:
source("example_fa.R")
```

## The Stan model

The key design choice is using the **marginal** likelihood. For a one-factor
model, this means integrating out the latent factor analytically:

$$\mathbf{y}_j \mid \boldsymbol{\theta} \sim \mathcal{N}_P(\boldsymbol{\alpha},\; \boldsymbol{\lambda}\boldsymbol{\lambda}^\prime + \mathrm{diag}(\boldsymbol{\sigma}^2))$$

In Stan:

```stan
transformed parameters {
  matrix[P, P] V;
  V = tcrossprod(to_matrix(lambda, P, 1)) + diag_matrix(square(sigma));
}

model {
  alpha ~ normal(0, 5);
  lambda ~ normal(0, 1);   // symmetric: allows sign switching
  sigma ~ normal(0, 2);

  for (j in 1:N) {
    Y[j] ~ multi_normal(alpha, V);
  }
}

generated quantities {
  vector[N] log_lik;
  for (j in 1:N) {
    log_lik[j] = multi_normal_lpdf(Y[j] | alpha, V);
  }
}
```

The `generated quantities` block outputs pointwise marginal log-likelihoods,
which is what `compute_dic_i()` expects.

## Simulate and fit

```{r simulate-fit}
library(cmdstanr)
library(DICi)

set.seed(2026)
N <- 400; P <- 6
lambda_true <- c(0.81, 0.72, 0.63, 0.54, 0.45, 0.36)
alpha_true  <- rep(0, P)
sigma_true  <- rep(sqrt(0.5), P)

eta <- rnorm(N)
Y <- matrix(NA, N, P)
for (j in 1:N) {
  Y[j, ] <- alpha_true + lambda_true * eta[j] + rnorm(P, 0, sigma_true)
}

mod <- cmdstan_model("fa_marginal.stan")
fit <- mod$sample(
  data = list(N = N, P = P, Y = Y),
  chains = 4, iter_warmup = 1000, iter_sampling = 1000,
  parallel_chains = 4, seed = 42
)
```

## Compute DIC_i

```{r dici}
# One-liner with the cmdstanr wrapper
result <- dic_i_from_cmdstanr(fit)
print(result)
# Parameterization-Invariant DIC (DIC_i)
# --------------------------------------
#   DIC_i  = 5724.55
#   p_V    = 17.74
#   E[D]   = 5706.81
# --------------------------------------
```

## Full comparison with all criteria

```{r compare-all}
library(loo)

log_lik <- fit$draws("log_lik", format = "draws_matrix")
log_lik <- unclass(log_lik)

# WAIC and LOO
waic_result <- waic(log_lik)
loo_result  <- loo(log_lik)

# Classical DIC
D_bar    <- mean(-2 * rowSums(log_lik))
D_plugin <- -2 * sum(colMeans(log_lik))
p_DIC    <- D_bar - D_plugin

cat(sprintf("DIC_i:         %8.1f  (p_V    = %5.1f)\n", result$dic_i, result$p_v))
cat(sprintf("WAIC:          %8.1f  (p_WAIC = %5.1f)\n",
            waic_result$estimates["waic", "Estimate"],
            waic_result$estimates["p_waic", "Estimate"]))
cat(sprintf("LOO-CV:        %8.1f  (p_LOO  = %5.1f)\n",
            loo_result$estimates["looic", "Estimate"],
            loo_result$estimates["p_loo", "Estimate"]))
cat(sprintf("DIC (classical): p_DIC = %.1f\n", p_DIC))
# Expected output:
#   DIC_i, WAIC, LOO-CV all agree closely (~5724-5725)
#   p_V, p_WAIC, p_LOO all near true k=18
#   p_DIC is negative (sign switching detected)
```

## Applying to your own models

The workflow generalizes to any model where you can compute marginal
log-likelihoods in Stan:

| Model class | How to get marginal log-lik |
|---|---|
| Factor analysis | `multi_normal_lpdf` with $\lambda\lambda^\prime + \Sigma$ |
| Linear mixed models | `multi_normal_lpdf` with $Z \Sigma Z^\prime + \sigma^2 I$ |
| Finite mixtures | `log_sum_exp` over component densities |
| Growth mixture models | Combine both: `log_sum_exp` over `multi_normal_lpdf` |

For models estimated in **JAGS** or **Mplus**, see `?compute_dic_i`.

## References

- Xiao, X. and Rabe-Hesketh, S. (2026). A Parameterization-Invariant DIC.
  *arXiv preprint* arXiv:2605.27844.
- Gelman, A., Hwang, J., and Vehtari, A. (2014). Understanding predictive
  information criteria for Bayesian models. *Statistics and Computing*,
  24, 997--1016.
- Merkle, E.C., Furr, D., and Rabe-Hesketh, S. (2019). Bayesian comparison
  of latent variable models: Conditional versus marginal likelihoods.
  *Psychometrika*, 84, 802--829.
